Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-04-06 FDAgroup.Rmd
edit introduction

df = content
df_train = subset(content, include == 1)
df_test = subset(content, include == 0)
```

@@

## Simulated child growth data

- Predict length-for-age z-score (LAZ), observed with noise
  - Calculated using the age- and sex-specific WHO standard references

```{r}
ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
  geom_line(show.legend = F)+
  labs(title = "Length-for-Age z-score tracks for the 10 subjects",
       x = "time", y ="LAZ")
```

## FPCA on observed LAZ

```{r}
tnew = sort(unique(df_train$argvals))
df_laz <- df_train %>% select(argvals, subj, Y) %>%
  rename("y" = "Y")
fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15,
                        argvals.new = tnew, calculate.scores = TRUE)

# mean function

par(mfrow = c(2, 2))
plot(tnew, fpca_laz$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")

```

## Prediction of new partially observed sample

```{r, pred_func}
# write a function for prediction up to t
# df_test: datafram with columns argvals, subj, y, observation up to tm

DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){
  Ntest = length(unique(df_test$subj))
  # prediction
  pred1 <- predict(fpca_fit, newdata = df_test)
  # Eigen functions
  Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ]
  Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores)
  # every column is a new function
  Yhat <- data.frame(Yhat)
  colnames(Yhat) <- pred1$rand_eff$subj
  df_pred <- Yhat %>% 
    mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>% 
    pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>%
    mutate(type = "pred")

  return(df_pred)
}
```

```{r t0_3}
df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.3))

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj == 152) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.3", 
       x = "time", y = "")
```

## Prediction of new partially observed sample

```{r}
df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.5), tm = 0.5)

# figure

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.5", 
       x = "time", y = "")

```

## Prediction of new partially observed sample

```{r}
df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y") %>% 
  filter(argvals <= 0.8), tm = 0.8)

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.8", 
       x = "time", y = "")

```

@@

```{r gen_bi_outcome}
times <- seq(0, 1, by = 0.001)
gen_data <- function(id = 1){
  scores <- rnorm(4)
  pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times)
  Z <- pcs %*% scores
  probs <- exp(Z)/(1+exp(Z))
  Y <- sapply(probs, rbinom, n = 1, size = 1)

  return(data.frame(id = id, argvals = times, Y=Y, Z=Z))
}

df_train2 <- bind_rows(lapply(1:20, gen_data))

ggplot(df_train2 %>% filter(id %in% 1:4))+
  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
            show.legend = F)+
  geom_line(aes(x=argvals, y=Z, col = "Red"), 
            show.legend = F)+
  facet_wrap(~id)+
  labs(title = "Simulated function for the 4 subjects",
       x = "time", y ="")
```

```{r}
# fPCA
fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15,
                        argvals.new = times, calculate.scores = TRUE)

# plot
par(mfrow = c(2, 2))
plot(times, fpca_Z$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")

```

>>>>

# df = content
# df_train = subset(content, include == 1)
# df_test = subset(content, include == 0)
```

@@

<!-- ## Simulated child growth data -->

<!-- - Predict length-for-age z-score (LAZ), observed with noise -->
<!--   - Calculated using the age- and sex-specific WHO standard references -->

<!-- ```{r} -->
<!-- ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+ -->
<!--   geom_line(show.legend = F)+ -->
<!--   labs(title = "Length-for-Age z-score tracks for the 10 subjects", -->
<!--        x = "time", y ="LAZ") -->
<!-- ```  -->

<!-- ## FPCA on observed LAZ -->

<!-- ```{r} -->
<!-- tnew = sort(unique(df_train$argvals)) -->
<!-- df_laz <- df_train %>% select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") -->
<!-- fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15, -->
<!--                         argvals.new = tnew, calculate.scores = TRUE) -->

<!-- # mean function -->

<!-- par(mfrow = c(2, 2)) -->
<!-- plot(tnew, fpca_laz$mu.new, type = "l",  -->
<!--      xlab = "t", ylab = "f0(t)") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2") -->
<!-- plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3") -->

<!-- ``` -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r, pred_func} -->
<!-- # write a function for prediction up to t -->
<!-- # df_test: datafram with columns argvals, subj, y, observation up to tm -->

<!-- DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){ -->
<!--   Ntest = length(unique(df_test$subj)) -->
<!--   # prediction -->
<!--   pred1 <- predict(fpca_fit, newdata = df_test) -->
<!--   # Eigen functions -->
<!--   Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ] -->
<!--   Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores) -->
<!--   # every column is a new function -->
<!--   Yhat <- data.frame(Yhat) -->
<!--   colnames(Yhat) <- pred1$rand_eff$subj -->
<!--   df_pred <- Yhat %>%  -->
<!--     mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>%  -->
<!--     pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>% -->
<!--     mutate(type = "pred") -->

<!--   return(df_pred) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r t0_3} -->
<!-- df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.3)) -->

<!-- # figure  -->
<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj == 152) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.3",  -->
<!--        x = "time", y = "") -->
<!-- ``` -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r} -->
<!-- df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.5), tm = 0.5) -->

<!-- # figure  -->

<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj %in% 151:156) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.5",  -->
<!--        x = "time", y = "") -->

<!-- ``` -->

<!-- ## Prediction of new partially observed sample -->

<!-- ```{r} -->
<!-- df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
<!--   rename("y" = "Y") %>%  -->
<!--   filter(argvals <= 0.8), tm = 0.8) -->

<!-- # figure  -->
<!-- bind_rows(df_test %>%  -->
<!--             select(argvals, subj, Y)%>%  -->
<!--             rename("y" = "Y") %>%  -->
<!--             mutate(type = "true", subj = as.factor(subj)),  -->
<!--           df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>%  -->
<!--   filter(subj %in% 151:156) %>%  -->
<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
<!--   geom_line()+ -->
<!--   facet_wrap(~subj)+ -->
<!--   labs(title = "Prediction with observation up to t=0.8",  -->
<!--        x = "time", y = "") -->

<!-- ``` -->

@@

<!-- ## Extension to non-Gaussian data -->

<!-- ```{r gen_bi_outcome} -->
<!-- times <- seq(0, 1, by = 0.001) -->
<!-- gen_data <- function(id = 1){ -->
<!--   scores <- rnorm(4) -->
<!--   pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times) -->
<!--   Z <- pcs %*% scores -->
<!--   probs <- exp(Z)/(1+exp(Z)) -->
<!--   Y <- sapply(probs, rbinom, n = 1, size = 1)  -->

<!--   return(data.frame(id = id, argvals = times, Y=Y, Z=Z)) -->
<!-- } -->

<!-- df_train2 <- bind_rows(lapply(1:20, gen_data)) -->

<!-- ggplot(df_train2 %>% filter(id %in% 1:4))+ -->
<!--   geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1, -->
<!--             show.legend = F)+ -->
<!--   geom_line(aes(x=argvals, y=Z, col = "Red"),  -->
<!--             show.legend = F)+ -->
<!--   facet_wrap(~id)+ -->
<!--   labs(title = "Simulated function for the 4 subjects", -->
<!--        x = "time", y ="") -->
<!-- ``` -->

<!-- ## Extension to non-Gaussian data -->

<!-- ```{r} -->
<!-- # fPCA -->
<!-- fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15, -->
<!--                         argvals.new = times, calculate.scores = TRUE) -->

<!-- # plot -->
<!-- par(mfrow = c(2, 2)) -->
<!-- plot(times, fpca_Z$mu.new, type = "l",  -->
<!--      xlab = "t", ylab = "f0(t)") -->
<!-- plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1") -->
<!-- plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2") -->
<!-- plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3") -->

<!-- ``` -->

### 2023-04-06 Write-ups/Manuscript_edit.Rmd
edit introduction

Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction. Traditionally, this type of data has been modeled either marginally, such as generalized estimating equations, or constitutionally on specific subjects, such as mixed effect models [@liang1986; @Laird1982; @lindstrom1990; @davidian2003]. For subject-specific models, predictions are made based on the correlation between repeated measures from the same subject and covariates that can be either fixed or varying. However, most of these methods are based on linear model structure, which is very limited in terms of flexibility of correlation. When sample size is large or the density of repeated measures is high, they also tend to cause severe computational burden [@GLMMadaptive]. In addition, with generalized, non-Gaussian outcomes, these methods lack the ability to handle out-of-sample prediction because there is usually no close-form solution for subject-specific random effects

To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures. Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function. Non-parametric smoothing [@Scheipl2014] are often incorporated to speed up the computation, where random effects are approximated by spline basis functions or eigenfunctions from functional principal component analysis (FPCA). The introduction of basis functions also makes out-of-sample prediction more straightforward. Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing. Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, usually modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors.

Unfortunately, fewer papers have focused on its extension to generalized, non-Gaussian outcomes, such as series of binary or count outcomes. Existing methods also tend to be very computationally intensive. For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes. @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*. @gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage joint estimation strategy. @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data. In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit GFPCA on binary data using EM algorithm, accompanied by the an open source R package *registr*.

In this paper, we aim to develop a fast, scalable method for dynamic prediction of generalized functional outcome based on functional mixed effect model with FPCA smoothing. Specifically, "Generalized" indicates that the outcome follows non-Gaussian exponential family distribution, and the term "functional" indicates that repeated measure are collected densely across the study domain (e.g., time). Section 2 explains the implementation procedure of the proposed method. In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study. In Section 4, we apply this method to a real world dataset from a weight loss study. Section 5 presents a discussion of advantages and limitation of the proposed method.

@@

where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process.

According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$. We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)$.

In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$. In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$. These eigenfunctions, also referred to as Principal Component (PC) functions, are orthogonal are can be orthonormal with proper scaling. $\xi_{ik}$ are subject-specific scores (or loadings) on each PC function. According to properties of FPCA method, these scores are mutually independent over both subject ($i$) and eigenfunctions ($k$). That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$, where $\lambda_k$ is the kth eigenvalues.

@@

3. Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, where $\epsilon_i(t)$ indicates random noise introduced by the binning process in previous steps. This random noise is assumed to follow a zero-mean Gaussian process, such that at each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$. The FPCA output includes estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.

@@

As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes. Just like many mixed models, this method is very limited in terms of flexibility. For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$. While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.

@@

The simulation study and case study of daily weigh-in data above have revealed the proposed method can achieve better predictive performance with much less computational burden. Compared to existing methods based on linear mixed model, fGFPCA can accommodate flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with wiggly or cyclic pattern. As in the simulation setting in Section 3, the binary outcomes are generated from a cyclic process. fGFPCA has much lower ISE and higher AUC than GLMM in this case, and its prediction accuracy also improves as more outcomes are collected, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data. Moreover, fGFPCA also significantly reduce time spent on model fitting and predict, which is beneficial when repeated measures are dense or sample size is large. In fact, as far as we are aware of, the proposed method is the only feasible method that can handle the scale of the simulated datasets with compatible flexibility.

@@

The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quantify such effects. Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid. Moreover, if the bins are very narrow with only a few observations, unidentifiability issues can happen when fitting local mixed models. While this may not necessarily affect point predictions, it can cause numeric failure to estimate standard error and prediction intervals. In practice, we expect this to happen when the measurements are sparse or sample size is small. Under such scenarios, one may choose traditional mixed model framework instead to get valid estimates at the expense of computational time.

It should also be noted that the binning procedure in Step 1 changes the density of repeated measures. We essentially assume that latent functions are smooth in small intervals, and treat adjacent observations as if they are generated from the same time point. The final prediction of latent continuous function tracks are made on the binned grid, integrating information from several adjacent observations in the same bin. In Section 3 and 4, this sparser grid was transformed to the original, un-binned grid using linear interpolation of the predicted latent function. However, this is by no means the only method to be considered. Such transformation can also be made by projection to a higher-dimensional space, similar as general additive model with a large number of basis functions. We suspect that this method would work better than linear interpolation when the original grid is sparse or irregular, but take longer time for computation.

>>>>

Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction. Traditionally, this type of data has been modeled either marginally, such as generalized estimating equations, or conditionally on specific subjects, such as mixed effect models [@liang1986; @Laird1982; @lindstrom1990; @davidian2003]. For subject-specific models, predictions are made based on the correlation between repeated measures from the same subject and covariates that can be either fixed or varying. Software and methods for dynamic prediction have been developed for Gaussian [leroux2016; @shang2017; @goldberg2014], non-Gaussian data [@chiou2012; @GLMMadaptive], and time-to-event outcomes [@suresh2017; @jmbook]. However, with the notable exception of @leroux2016, existing methods are either limited in their flexibility (both fixed and random effect structure, e.g. random slope and intercept), computationally unfeasible for large datasets, or both. These issues are particularly acute for data which are measured densely, as is often the case with functional data, or when data are non-Gaussian.

While methods with accompanying software for modeling complex functional data with non-linear fixed and random effects exist for both Gaussian and non-Gaussian data, software is largely limited to the refund package in the statistical software R [@Scheipl2014; @wood2014; @refundpkg]. However, these methods and software implementations do not currently allow for dynamic prediction. In addition to the methods implemented in the refund package, other work has been done on the fitting of functional regression models to non-Gaussian functional data. For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes. @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*. @gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage joint estimation strategy. @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data. In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit GFPCA on binary data using EM algorithm, accompanied by the an open source R package *registr*.

Existing research on dynamic prediction with functional data has focused primarily on continuous/Gaussian outcomes, usually modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors. Although methodological development exists, flexible software for implementing these methods is still relatively sparse with the exception of the fcr [@fcrpkg] package accompanying [@leroux2016].

In this paper, we aim to develop a fast, scalable method for dynamic prediction of generalized functional outcomes using the principals of local generalized linear mixed effects models combined with functional principal components analysis of the estimated latent processes. The proposed method is designed to handle dense, regularly measured functional data and is extremely computationally efficient relative to the model complexity. Section 2 explains the implementation procedure of the proposed method. In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study. In Section 4, we apply this method to a real world dataset from a weight loss study. Section 5 presents a discussion of advantages and limitation of the proposed method.

@@

where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process with covariance operator $\Sigma_b$. By the Karhunen-Loève theorem, $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$. In practice, $b_i(t)$ is approximated using a finite $K$ number of basis functions.

Estimation of $b_i(t)$ involves estimating both $\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}$ and $\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}$. For Gaussian data this may be done by directly estimating the covariance operator $\Sigma_b$ using either method of moments (if the data are relatively small) or using faster methods such as the FACE method [reference - FACE]. However, for non-Gaussian data, direct estimation is not possible. Instead, one may choose to estimate $b_i(t)$ first using a flexible spline basis  (e.g. B-splines)  with dimension $K^b$ such that $K^B > K$, assuming an unstructured correlation between subject-specific spline coefficients, then orthogonalizing the resulting estimates via eigendecomposition or singular value decomposition. This general approach forms the basis for estimation for several GFPCA methods [references - goldsmith, julia's EM algorithm]. However, this approach results in rapidly escalating computational burden as both the number of subjects, $N$, and the density of the observed functional predictor increases. Note that if the covariance operator $\Sigma_b$ is relatively complex, then $K$ may be large and not well approximated using simple basis functions, $\phi_k$ (e.g. random intercept and slope).

Beyond the computational burden associated with fitting the model of interest, for dynamic prediction one must then estimate the scores for a new subject $u$ not used in model fitting. Typically this takes the form of the conditional expectation of the scores $E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]$ where $\bf{\Theta}$ represents the set of parameters estimated by the model (i.e. parameters associated with $\beta_0(t), \Sigma_b$) and $\bf{y}_u$ are the non-Gaussian data for the new subject. Estimation of the random effects for a new subject is non-trivial and not implemented for even very simple models in most random effects software packages such as lme4 [reference] and nlme [reference]. Our proposed method bypasses this problem by using local generalized linear mixed models to estimate the covariance operator $\Sigma_b$ directly on the latent scale, then using the closed form solution for the best linear unbiased predictor of Gaussian random effects. Estimation is made fast for even very large data through the use of the FACE methodology for FPCA on the latent estimates obtained from local models. Because the local models are very simple GLMMs, estimation is also fast for the local models.

@@

3. Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.

**Ying, include the formulas for your conditional expectation here to obtain random effects estimates **

@@

As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes. Just like many mixed models, this method is very limited in terms of flexibility if one wishes to obtain estimates in a reasonable time frame. For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$. While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.

@@

The simulation study and data application above have illustrated the feasibility and utility of the proposed method, even for large datasets and complex latent processes/random effect structures. can achieve better predictive performance with much less computational burden. Compared to competing methods, fGFPCA can accommodate extremely flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with highly non-linear patterns. Due to the comparative flexibility of the fGFPCA approach, fGFPCA has much lower ISE and higher AUC than GLMM. As expected, the dynamic prediction accuracy of fGFPCA improves as more data are availble for predictions. Moreover, fGFPCA also significantly reduce time spent on both model fitting and prediction, allowing for scaling of models and predictions which are infeasible with existing approaches. In fact, as far as we are aware of, the proposed method is the only feasible method that can handle the scale of the simulated datasets with compatible flexibility.

**Andrew: I would delete this paragraph, honestly**

@@

A few open questions remain with the fGFPCA approach. The choices involved in the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quantify such effects. Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid. Moreover, if the bins are very narrow with only a few observations, local models may be non-identifiabile, resulting in bins with no point estimates. While this may not overly impact point final predictions, it is currently unclear as to when this issue warrants concern. In practice, we expect this to happen when the measurements are sparse or sample size is small. However, the methodology here is designed for large, dense functional data and thus not currently appropriate in such scenarios. In this case, existing methods may be sufficiently flexible and computationally feasible.
